Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression

### 2020-05-19 paper/synthetic.tex
synthetic

We evaluate the performance of B2B throughout a series of experiments on
controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2B in terms of
prediction of independent and identically distributed data, as well as a method
to recover causal factors.

@@

We compare the performance of B2B against four competing methods, all
implemented in scikit-learn \citep{sklearn} and pyrcca \citep{bilenko2016pyrcca}:
%
% To be updated

@@

@@

B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
features. In contrast, the Forward, Backward, CCA and PLS models lead to a
loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
model). To transform such vector into an estimated causal contribution $\hat E$,
we take the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}

To estimate whether models accurately identify causal factors, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
The AUC allows evaluating the capacity of models at detecting the causal
importance of factors when ground truth labels are available, as is the case in
this setup.

@@

In most cases, $E$ is not known and AUC can thus not be estimated. To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-ones feature $X_{-i}$ (i.e. 'knock-out X'). This procedure
results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each $X_i$
improves the prediction of $Y$.
%beyond what can be predicted from all other features $X_{-i}$.
In our figures, $\Delta R$ is the average of $\Delta R_i$. A higher score means that for prediction, the model relies on individual features rather than combinations of features.

We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines. For the Backward Model, feature importance cannot be assessed as the $X$ collinearity is never taken into account.

% Feature importance can be computed similarly for the CCA, PLS and Forward baselines (see Appendix \ref{appendix:feature_importance}). For the Backward Model, feature importance cannot be assessed as there is no prediction.

We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method outperforms baselines.

%
% \iffalse
% For CCA and PLS, feature importance is assessed as follow:
% \begin{enumerate}
%   \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
%   \item For each feature $i$:
%   \begin{itemize}
%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
%     \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
%   \end{itemize}
% \end{enumerate}
% \fi
% For the Backward Model, feature importance cannot be assessed because there is no prediction.
% We thus simply
% report the model performance $R_i = corr(YG_i, X_i)$.

% \iffalse
% \begin{enumerate}
%
%   \item Fit $G$ given $X_{train}$ and $Y_{train}$
%   \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
%   \item For each feature $i$:
%   \begin{itemize}
%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
%   \end{itemize}
%
%   \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
%   \item For all $i \in d_x$:
%   \begin{itemize}
%     \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
%     \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
%   \end{itemize}
%   \end{enumerate}
% \fi

% \iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
% \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
% tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
% Reduced Rank Regression or RRR \citep{Izenman_rrr}.
% %
% Each of these methods estimates a matrix of coefficients $\hat{W} \in
% \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
% using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
% %
% For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
% %
%
%
% We evaluate seven metrics:
% %
% test error in-distribution, test error out-distribution, test error-in
% distribution after weighting with $\hat{E}$, test error out-distribution after
% weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
% Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
% (AUC) between $\hat{E}$ and $E$.
% %
% The ``in-distribution'' metrics measure the test error of each predictor before
% and after weighting their features by the estimated (continuous!) mask
% $\hat{E}$.
% %
% The ``out-distribution'' metrics are similar, measuring the test error at
% samples drawn using two different random covariance matrices for both $X$ and
% $N$.
% %
% The false positives, false negatives, and AUC statistics measure the quality of
% identification of causal factors between $\hat{E}$ and the true, binary $E$.
%
% \begin{figure}[htpb] \centering
% \includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
% experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
% against a competitor ($y$-axis) for each experiment configuration averaged over
% $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
% the training distribution are weakly but consistently larger for all of the
% models as compared to B2B with and without the Sonquist binarization (E). The
% AUC, measuring the ability of each model to separate causal from noncausal
% factors, is consistently larger in B2B as compared to the other models.}
% \label{table:synthetic} \end{figure} \fi

% Figure~\ref{table:synthetic} summarizes the results for all experimental
% conditions, metrics, and methods.
% %
% B2B is the method obtaining best test errors, lowest false positive and false
% negative discovery about $E$, and overall the best AUC when it comes to
% detecting causal influences.

>>>>

We evaluate the performance of B2B throughout a series of experiments on
controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2B on its
ability to 1) accurately predict independent and identically distributed data,
and 2) recover causal factors when the ground truth is known.

@@

We compare the performance of B2B against four baseline methods.
%
% To be updated

@@

All baseline method were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
\citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.

@@

%
B2B leads to \emph{scalar} coefficients for non-causal
features. The diagonal of this matrix, $\hat E$, can thus be directly used
as a causal contribution estimate. Note that is estimate is unbiased (i.e.
zeros-centered) if the second regression $H$ is not regularized.

In contrast, the Forward, Backward, CCA and PLS models lead to a
loading \emph{vector} for each feature: each vector indicates how to it
relates to each dimension of the observations (for forward or backward
models) or each canonical projection of the observations
To an estimated causal contribution, we must thus transform such vector into a
scalar, by e.g. taking the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}
Note that in such B2B cases, the estimates are necessarily biased: i.e. positive.

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup.

@@

In most cases, $E$ is not known. The above AUC can thus not be estimated.
To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
each feature (for the B2B and Backward models) for each dimension of $Y$ (for
the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.

We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess
feature importance for each model. For the Backward Model, feature importance
cannot be assessed as the $X$ collinearity is never
taken into account.

In our synthetic figures, we do report each feature $i$ separately, as they are
randomly generated. $\Delta R$ thus represents the average of $\Delta R_i$. A
higher score means that for prediction, the model relies on individual features
rather than combinations of features.

Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
Appendix) show that B2B outperforms baseline models.
